Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} Stagnating State Funding for Higher Education and its Effect on Faculty

### 2024-09-26 text/sections/01-introduction.tex
Made some edits.
Added a number of new things.  2 minor comments to address before sub.

Per-student state funding for higher education has stagnated the last thirty years.
The average public university received around \$11,600 of state funding per-student in 1990 and \$8,300 in 2021, while the number of lecturers per-student increased by 109\% and professors per-student decreased by $23$\% over 1990--2021.
I use a shift-share instrumental variables (IV) approach to estimate how faculty are affected by these funding cuts.
Universities employ $4.4$\% more lecturers per-student following a $10$\% funding cut, and correspondingly rely less on tenure-track faculty by employing $1.4$\% fewer tenure-track professors and $1.2$\% fewer tenured professors per-student.
In a secondary analysis of all public university faculty in the state of Illinois, I show that incumbent faculty were unaffected, in terms of wages, promotion and quit rates.
This implies that the faculty substitution arose by disrupting hiring of new tenure-track/tenured professors, which is supported supplementary data on professor hiring at US public universities over 2010--2021.
This paper shows the institutional effects of stagnating state funding for higher education, and offers one possible mechanism for how funding cuts affected student outcomes.

@@

I provide the first causal evidence on the impact of state funding on faculty composition and wages.
In a contemporaneous working paper, \cite{hinrichs2022state} use the same national-level variation in university funding to analyse effects on faculty and wages, coming to different conclusions.\footnote{    
    \cite{hinrichs2022state} considers a shorter time-period than this paper (2010--2021), uses only national variation in IPEDS data, relies on IPEDS data for faculty salaries, and has no individual-level analysis of faculty or their salaries.
    See \autoref{sec:data-desc,sec:recession-analysis} for an elaboration on these issues.
}
The closest related research on higher education funding examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,webber2017state}.
For faculty outcomes, descriptive work measures how US faculty salaries are correlated with other factors \citep{hilmer2020labor}, and applied theory work models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
Other papers measure university decision--making after endowment shocks \citep{brown2014endowment}, and trends in the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure changes in state funding and university revenues, but do not measure effects of state funding cuts on faculty composition or salaries, or other outcomes related to university instruction.
My results provide evidence that the long term trends in higher education funding and faculty outcomes are causally related, contributing to the literature on faculty outcomes \citep{ehrenberg2003studying} and trends in US higher education and funding \citep{hoxby2009changing,ehrenberg2012american}.

>>>>

Per-student state funding for higher education has stagnated the last thirty years.
The average public university received around \$11,600 of state funding per-student in 1990 and \$8,300 in 2021, while the number of lecturers per-student increased by 109\% and professors per-student decreased by $23$\% over 1990--2021.
I use a shift-share instrumental variables (IV) approach to estimate how faculty are affected by these funding cuts.
Universities employ $4.4$\% more lecturers per-student following a $10$\% funding cut, and correspondingly rely less on tenure-track faculty by employing $1.4$\% fewer tenure-track professors and $1.2$\% fewer tenured professors per-student.
In a secondary analysis of all public university faculty in the state of Illinois, I show that incumbent faculty were unaffected, in terms of wages, promotion and quit rates.
This implies that the faculty substitution arose by disrupting hiring of new tenure-track/tenured professors, which is supported with supplementary data on professor hiring at US public universities over 2010--2021.
This paper shows the institutional effects of stagnating state funding for higher education, and offers one mechanism for how funding cuts affected student outcomes.

@@

I provide the first causal evidence on the impact of state funding on faculty composition and wages.
In a contemporaneous working paper, \cite{hinrichs2022state} use the same national-level variation in university funding to analyse effects on faculty and wages, coming to different conclusions.\footnote{
    An anonymous reviewer made me aware of a similar and contemporary working paper, \cite{hinrichs2022state}; this work analyses a shorter time-period than this paper, and only focuses on national variation.
    \autoref{sec:recession-analysis} replicates their key findings, and shows how their focusing on only the 2012--2019 years explains differences in conclusions.
}
The closest related research on higher education funding examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,webber2017state}.
For faculty outcomes, descriptive work measures how US faculty salaries are correlated with other factors \citep{hilmer2020labor}, and applied theory work models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
Other papers measure university decision--making after endowment shocks \citep{brown2014endowment}, and trends in the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure changes in state funding and university revenues, but do not measure effects of state funding cuts on faculty composition or salaries, or other outcomes related to university instruction.
My results provide evidence that the long term trends in higher education funding and faculty outcomes are causally related, contributing to the literature on faculty outcomes \citep{ehrenberg2003studying} and trends in US higher education and funding \citep{hoxby2009changing,ehrenberg2012american}.

### 2024-09-26 text/sections/04-empirical.tex
Added a number of new things.  2 minor comments to address before sub.

The system exploits the fact that institutions who rely more on state funding will be affected by follow-on state-wide funding cuts.
It is constructed by interacting the average funding for universities in state $s(i)$ with reliance on state funding relative to total revenues, averaged across the base years 1990--1993.\footnote{
    1990--1993 are defined as data for public university finance data are most comparable (i.e., without many missing values) beginning in 1990.
    \cite{NBERw23736} use the single year 1990 as the base year, though I use the four years to ameliorate missing values in the single year of 1990.
    Results are similar in either specification.
}
$Z_{i,t}$ is negative to reflect the fact that the long term trend in, and most of the short-run shifts in, state funding for higher education have been negative.\footnote{
    When used in log terms, the instrument is the negative of the logged shock --- i.e., $- \log \left( -Z_{i,t} \right)$ as opposed to log of the negative shock $\log Z_{i,t}$ directly.
}
State funding has been falling, so that the instrument describes shocks to university revenues, mostly in a negative direction.\footnote{
    \label{foot:control}
    \cite{NBERw27885,webber2017state} note the tendency for public universities to respond to state funding cuts by increasing reliance on tuition, where \cite{NBERw23736} specifically instruments for tuition revenues with collected information on legislative tuition price controls.
    It may be argued that tuition revenues are confounder between the causal effect of changes in state funding on a university's total revenues, so that this analysis focuses on state support for higher education (not total revenue), as do \cite{NBERw27885}.
    On the other hand, rises in tuition revenues (per-student) may arise as result of tuition hikes thanks declining state support, which would mean controlling for tuition would constitute a bad control.
    Nonetheless, estimates including tuition revenues (per-student) as a control in the second stage of the IV estimates produce results of very similar magnitude and direction, and so are omitted.
}

>>>>

The system exploits the fact that institutions who rely more on state funding will be affected by follow-on state-wide funding cuts.
It is constructed by interacting the average funding for universities in state $s(i)$ with reliance on state funding relative to total revenues, averaged across the base years 1990--1993.\footnote{
    1990--1993 are defined as data for public university finance data are most comparable (i.e., without many missing values) beginning in 1990.
    \cite{NBERw23736} use the single year 1990 as the base year, though I use the four years to ameliorate missing values in the single year of 1990.
    Results are similar in either specification.
}
$Z_{i,t}$ is negative to reflect the fact that the long term trend in, and most of the short-run shifts in, state funding for higher education have been negative.\footnote{
    When used in log terms, the instrument is the negative of the logged shock --- i.e., $- \log \left( -Z_{i,t} \right)$ as opposed to log of the negative shock $\log Z_{i,t}$ directly.
}
State funding has been falling, so that the instrument describes shocks to university revenues, mostly in a negative direction.\footnote{
    \label{foot:control}
    \cite{NBERw27885,webber2017state} note the tendency for public universities to respond to state funding cuts by increasing reliance on tuition, where \cite{NBERw23736} specifically instruments for tuition revenues with collected information on legislative tuition price controls.
    It may be argued that tuition revenues are confounder between the causal effect of changes in state funding on a university's total revenues, so that this analysis focuses on state support for higher education (not total revenue), as do \cite{NBERw27885}.
    On the other hand, rises in tuition revenues (per-student) may arise as result of tuition hikes following state funding cuts, which would mean controlling for tuition constitutes a bad control.
    Nonetheless, estimates including tuition revenues (per-student) as a control in the second stage of the IV estimates produce results of very similar magnitude and direction, and so are omitted (see \autoref{sec:results-firststage}).
}

### 2024-09-26 text/sections/05-results.tex
Added a number of new things.  2 minor comments to address before sub.

\subsection{First-Stage Estimates}
The shift-share instrument has a strong first-stage for universities' yearly state funding.
\autoref{tab:firststage-reg} show results of the first-stage regression among IPEDS data, separately with and without a control for tuition revenue per student.
\autoref{tab:firststage-reg} Panel A shows estimates of a state funding shift-share of \$1 per student on state funding per student at the university, and Panel B the effects of a 1\% increase in the state funding shift-share on state funding per student.
Column (1) shows that a state funding shift-share of \$10 per student is associated with \$11.76 less state funding per student at the university; $-10$\% state funding shift-share leads to $-9.77$\% state funding per student at the average university.
Estimates are presented for specifications with and without fixed effects.
Column (2) shows estimates of the first-stage without including fixed effects, and gives less precise estimates for the funding shift-share --- this may be due to systematic differences in universities, which are not accounted for when omitting fixed effects.
Columns (3) and (4) include the tuition revenue as a control (explained in \autoref{foot:control}) to exhibit estimates with or without this control variable.
Column (3) shows similar estimates to column (1) in both Panels A and B thanks to inclusion of fixed effects, so whether tuition revenue is a bad control does not affect empirical estimates.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects (not including tuition revenue as a control variable), and is the preferred specification that I proceed with;
all specifications (except column 4 in both panels) have a large F statistic on the excluded instrument.\footnote{
    I calculate the F statistic on the excluded instrument, while also accounting for fixed effects and two-way standard error clustering, by the methods provided in \cite{olea2013robust}.
}
I use the same empirical strategy in the analysis of Illinois faculty, and \autoref{tab:firststage-illinois} presents first-stage estimates among the IBHED data; results are similar to those calculated in IPEDS data.
These results show the funding shift-share instrument strongly predicts a university's state funding in the first-stage estimation.

@@

Do state funding cuts meaningfully affect universities?
While the shift-share performs well in predicting state funding (i.e., a strong instrument), it could be the case that the day-to-day operations of a university are not affected.
University could have responded to state funding cuts by cutting areas of spending unrelated to research or instruction, or fully off-set the cuts by charging higher tuition fees, leading to no effects on the university or faculty.
To allay this concern, I estimate the IV model on IPEDS data for areas of university spending  (\autoref{tab:expenditures-shock-reg}).
This exercise gauges the effects of state funding cuts on areas of university spending to gauge whether the first-stage is economically meaningful, and to see which areas of higher education operations are affected by state funding cuts.
Following a state funding cut, universities spend less on instruction, student services, operations, and grant aid.
Additionally, they charge higher tuition fees, though this rise is at most a quarter of the state funding cut, so does not come close to fully off-setting state funding cuts.
These results are in line with previous work showing state funding cuts (or tuition price limits) reduce student-focused spending \citep{NBERw23736}, financial aid \citep{miller2022making}, and lead to increased tuition fees \citep{bound2019public}.

@@

\subsection{Robustness Checks and Heterogeneity Analysis}
\label{sec:results-robustness}
It is possible that the identification strategy is not robust to some confounding issues or trends.
For example, selective institutions may have received funding cuts for other reasons, or states with lower performing institutions or declining populations strategically cut funding for only less selective universities.
Additionally, the effect of state funding cuts may affect faculty at different universities to different extents
I investigate these issues with three different analyses.

@@

Second, I return to the primary specification in \autoref{eqn:secondstage}, and estimate effect of state funding cuts on faculty counts separately among universities of different selectivity rates.
This investigates whether substitution rates are different across different levels of selectivity; \autoref{tab:facultycount-heterogeneity} collects these results together, separated by the \cite{barrons2009} selectivity rankings.
Barrons rank 1-2 are the most selective (including e.g., UC Berkeley and SUNY Binghamton), 3-4 selective (including e.g., Clemson University and University of Dallas), and rank 5 for the rest of universities.
The first-stage is strong in every group of universities, even among the small sample size of the 12 universities in the most selective group, and the 28 universities in the next selective group.
Faculty substitution rates are not very strong among the most selective universities, but are very similar between selective universities and unranked universities.
One might expect that the effect is driven entirely by the least selective universities, as they are the most resource-constrained;
this heterogeneity analysis shows that even selective (if not top) universities substitute towards lecturers following state funding cuts.

The final robustness check I employ is a falsification test, testing whether the effects for public universities would be picked up among a sample of private universities.\footnote{
    This is a robustness check similar to one performed in \citealt{NBERw27885}, as a robustness check for the same instrument.
}
Private universities do not rely on state funding in the same way that public universities do,\footnote{
    Though private universities do receive some state funding, as some states fund smaller scale operating costs and other specific projects at private universities.
}
so we would expect state funding to not affect their faculty counts.
If the state funding and faculty counts were related among private universities, this would be evidence that the IV model picks up otherwise concurrent trends in higher education, unrelated to stagnating state funding for public universities.
To impute a base share of reliance on state funding, I assign private universities the average reliance on state funding in the base period of universities in the same state with the same Barrons selectivity ranking, and then compute the shift-share instrument from this value.
\autoref{tab:facultycount-shock-private-robustness} shows that the first-stage is not strong among private universities; despite this, the IV results are still not significant for any faculty count, showing that state funding is not related to private universities' faculty counts.
The OLS columns show that imputed state funding is highly correlated with lecturer and professor counts at private universities.
The IV estimates are not distinguishable from zero, showing that OLS correlations are not causal relationships, and validating the IV approach for inference on universities' faculty counts.

>>>>

\subsection{First-Stage Estimates}
\label{sec:results-firststage}
The shift-share instrument has a strong first-stage for universities' yearly state funding.
\autoref{tab:firststage-reg} show results of the first-stage regression among IPEDS data, separately with and without a control for tuition revenue per student.
\autoref{tab:firststage-reg} Panel A shows estimates of a state funding shift-share of \$1 per student on state funding per student at the university, and Panel B the effects of a 1\% increase in the state funding shift-share on state funding per student.
Column (1) shows that a state funding shift-share of \$10 per student is associated with \$11.76 less state funding per student at the university; $-10$\% state funding shift-share leads to $-9.77$\% state funding per student at the average university.
Estimates are presented for specifications with and without fixed effects.
Column (2) shows estimates of the first-stage without including fixed effects, and gives less precise estimates for the funding shift-share --- this may be due to systematic differences in universities, which are not accounted for when omitting fixed effects.
Columns (3) and (4) include the tuition revenue as a control (explained in \autoref{foot:control}) to exhibit estimates with or without this control variable.
Column (3) shows similar estimates to column (1) in both Panels A and B thanks to inclusion of fixed effects, so whether tuition revenue is a bad control does not affect empirical estimates.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects (not including tuition revenue as a control variable), and is the preferred specification that I proceed with;
all specifications (except column 4 in both panels) have a large F statistic on the excluded instrument.\footnote{
    I calculate the F statistic on the excluded instrument, while also accounting for fixed effects and two-way standard error clustering, by the methods provided in \cite{olea2013robust}.
}
I use the same empirical strategy in the analysis of Illinois faculty, and \autoref{tab:firststage-illinois} presents first-stage estimates among the IBHED data; results are similar to those calculated in IPEDS data.
These results show the funding shift-share instrument strongly predicts a university's state funding in the first-stage estimation.

@@

Do state funding cuts meaningfully affect universities?
While the shift-share performs well in predicting state funding (i.e., a strong instrument), it could be the case that the day-to-day operations of a university are not affected.
University could have responded to state funding cuts by cutting areas of spending unrelated to research or instruction, or fully off-set the cuts by charging higher tuition fees, leading to no effects on the university or faculty.
To allay this concern, I estimate the IV model on IPEDS data for areas of university spending  (\autoref{tab:expenditures-shock-reg}).
This exercise gauges the effects of state funding cuts on areas of university spending to gauge whether the first-stage is economically meaningful, and to see which areas of higher education operations are affected by state funding cuts.
Following a state funding cut, universities spend less on instruction, student services, operations, and grant aid.
Additionally, they charge higher tuition fees, though this rise is at most a quarter of the state funding cut, so does not come close to fully off-setting state funding cuts.
These results are in line with previous work showing state funding cuts (or tuition price limits) reduce student-focused spending \citep{NBERw23736}, financial aid \citep{miller2022making}, and lead to increased tuition fees \citep{bound2019public,webber2017state}.

@@

\subsection{Robustness Checks and Heterogeneity Analysis}
\label{sec:results-robustness}
It is possible that the identification strategy is not robust to some confounding issues or trends.
For example, selective institutions may have received funding cuts for other reasons, or states with lower performing institutions or declining populations strategically cut funding for only less selective universities.
I investigate these issues with two different analyses, and then investigate heterogeneity across rates of selectivity and in different decades.

@@

As another robustness check, I employ is a falsification test, testing whether the effects for public universities would be picked up among a sample of private universities.\footnote{
    This is a robustness check similar to one performed in \citealt{NBERw27885}, as a robustness check for the same instrument.
}
Private universities do not rely on state funding in the same way that public universities do,\footnote{
    Some private universities do receive state funding, mostly funding for smaller scale operating costs and other specific projects.
}
so we would expect state funding to not affect their faculty counts.
If the state funding and faculty counts were related among private universities, this would be evidence that the IV model picks up otherwise concurrent trends in higher education, unrelated to stagnating state funding for public universities.
To impute a base share of reliance on state funding, I assign private universities the average reliance on state funding in the base period of universities in the same state with the same Barrons selectivity ranking, and then compute the shift-share instrument from this value.
\autoref{tab:facultycount-shock-private-robustness} shows that the first-stage is not strong among private universities; despite this, the IV results are still not significant for any faculty count, showing that state funding is not related to private universities' faculty counts.
The OLS columns show that imputed state funding is highly correlated with lecturer and professor counts at private universities.
The IV estimates are not distinguishable from zero, showing that OLS correlations are not causal relationships, and validating the IV approach for inference on universities' faculty counts.

To investigate whether universities of different selectivity respond differently to state funding cuts, I estimate the effect of state funding cuts on faculty counts separately among universities of different selectivity rates.
\autoref{tab:facultycount-heterogeneity} collects these results together, estimating \autoref{eqn:secondstage} separately by the \cite{barrons2009} selectivity rankings.
Barrons rank 1-2 are the most selective (including e.g., UC Berkeley and SUNY Binghamton), 3-4 medium selective (including e.g., Clemson University and University of Dallas), and rank 5 for the rest of universities.
The first-stage is strong in every group of universities, even among the small sample size of 12 universities in the most selective group, and 28 universities in the next selective group.
Faculty substitution rates are not distinguishable from zero among the most selective universities, but are positive and similar between both medium selective universities and unranked universities.
One might expect that the effect is driven entirely by non-selective and low ranked universities, as they are the most resource-constrained;
this heterogeneity analysis shows that even medium selective universities substitute towards lecturers following state funding cuts.

Last, I investigate when the substitution towards lecturers occurred.
To do so, I estimate the effect of state funding on faculty counts relative student enrolment separately for 1990--2009 and 2010--2021.
Substitution towards lecturer is weak to non existent in the latest decade, as noted in contemporary research \citep{hinrichs2022state}.
However, the rates are stronger when just looking at the years 1990--2010.
This reflects a key point noted in \autoref{sec:trends}: a large amount of the cuts, and general stagnation, in state funding for public universities occurred over in the 1990s and 2000s.
Additionally, the trend in rising lecturer count, and falling full professor count, per student at public universities had been operating all throughout the 1990s and 2000s.
These trends did continue in the 2010s (especially for counts of assistant professors), but most of the substitution had already taken place by this point.

### 2024-09-26 text/sections/08-figures-tables.tex
Made some edits.
Added a number of new things.  2 minor comments to address before sub.

\begin{table}[H]
    \singlespacing
    \centering
    \caption{First Stage Estimates, Predicting State Funding by Funding Shift-Share, in IPEDS Data.}
    \textbf{Panel A: units in \$ per student}

@@

    \label{tab:firststage-reg}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the first stage OLS estimates of regression specification \eqref{eqn:firststage}, showing the effect of the state funding shift-share on state funding to gauge performance as an instrument.
    Each observation is a public university-year, in the IPEDS data.
    Panel A shows the effect of a state funding shift-share of \$-1 per student in the state on the number of \$'s of state funding per student at the university --- i.e.,
    \$-1 state funding shift-share per student in the state leads to \$1.176 less state funding per student at the university according to preferred specification column 1.
    Panel B shows the effect of a $-10$\% change state funding shift-share per student in the state on $10$\% change in state funding per student at the university --- i.e.,
    $-10$\% state funding shift-share per student in the state leads to $-9.77$\% less state funding per student at the university according to prefferred specification column 1.        
    Standard errors are clustered at the state-year level, and university $+$ year fixed effects are included where noted.
\end{table}

>>>>

\begin{table}[H]
    \singlespacing
    \centering
    \caption{First-Stage Estimates, Predicting State Funding by Funding Shift-Share, in IPEDS Data.}
    \textbf{Panel A: units in \$ per student}

@@

    \label{tab:firststage-reg}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the first-stage OLS estimates of regression specification \eqref{eqn:firststage}, showing the effect of the state funding shift-share on state funding to gauge performance as an instrument.
    Each observation is a public university-year, in the IPEDS data.
    Panel A shows the effect of a state funding shift-share of \$-1 per student in the state on the number of \$'s of state funding per student at the university --- i.e.,
    \$-1 state funding shift-share per student in the state leads to \$1.176 less state funding per student at the university according to preferred specification column 1.
    Panel B shows the effect of a $-10$\% change state funding shift-share per student in the state on $10$\% change in state funding per student at the university --- i.e.,
    $-10$\% state funding shift-share per student in the state leads to $-9.77$\% less state funding per student at the university according to prefferred specification column 1.        
    Standard errors are clustered at the state-year level, and university $+$ year fixed effects are included where noted.
    The F statistic is for the excluded instrument (not for overall regression), after accounting for fixed effects and cluster dependence \citep{olea2013robust}.
\end{table}

### 2024-09-26 text/sections/09-appendix.tex
Added a number of new things.  2 minor comments to address before sub.

\newpage
\subsection{Analysis of IPEDS Salary Data and Years Following the Great Recession}
\label{sec:recession-analysis}

Bigger analysis in here.

>>>>

\newpage
\subsection{Different Rates of Substitution for Each Decade}
\label{sec:decade-analysis}

\begin{table}[H]
    \singlespacing
    \centering
    \caption{Effects of State Funding on Faculty Counts, in IPEDS Data 1990--2021, Separated for Each Decade.}

    \textbf{Panel A: Estimated Separately for 1990--2009}

    \makebox[\textwidth][c]{
        \input{tables/facultycount-replicated-corrected-1990s.tex}}

    \textbf{Panel B: Estimated Separately for 2010--2021}

    \makebox[\textwidth][c]{
            \input{tables/facultycount-replicated-corrected-2010s.tex}}

    \label{tab:facultycount-replicated-corrected}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the second stage OLS and 2SLS estimates of regression specification \eqref{eqn:secondstage}, showing the effect of state funding changes on faculty outcomes, using the state funding shift-share (based in 1990--1993) to instrument for state funding in the columns labelled 2SLS.
    Each panel gives estimates for data restricted to the noted decades.
    Each observation is a public university-year, in the IPEDS data.
    Each panel shows the effect of a $10$\% change in state funding per student at the university on the 10\% change in the number of professors per students.
    Outcome-mean is the mean of the outcome, for Panel A the number of professors per student, for Panel B the number of faculty per student.
    Each uses $\log$ faculty count per student as the outcome, though the outcome mean is count of faculty per student (not in $\log$ terms).
    Standard errors are clustered at the state-year level, and university $+$ year fixed effects are included through-out.
\end{table}

\newpage
\subsection{Replication of IPEDS Results for the Years 2011--2019}
\label{sec:recession-analysis}
\cite{hinrichs2022state} present an analysis of the same topic as the IPEDS analysis in this paper.
Their analysis also includes faculty at 2-year colleges, which is not related to my analysis of lecturers and professors at 4-year universities.

\cite{hinrichs2022state} analyses IPEDS data only in the years 2011--2019, as the Department of Education added a number of questions to the IPEDS survey asking how many part- and full-time professors are employed by the university.
This means that they define the shift-share instrument with a base year of 2011, do not adjust for the student population size --- in neither the state nor the university under consideration (i.e., on neither shift nor share):
\begin{align}
    \label{eqn:replication-instrument}
    Z^{2011}_{i,t} &=
    \text{Total State Funding}_{s(i),t} \times
    \left( \frac{\text{State Funding}_{i,2011}}{\text{Total Revenues}_{i,2011}} \right).
\end{align}
I replicate the $Z^{2011}_{i,t}$ formulation of the shift-share instrument, showing it is strong in predicting state funding (without adjusting for student population) in following years 2012--2019.
There are only minor differences in the results presented to that for Table 2 in \cite{hinrichs2022state}, as their estimation is on the sample of universities across different samples.
\autoref{tab:firststage-replicate} replicates this finding on the sample of universities who report faculty counts by rank.
\begin{table}[H]
    \singlespacing
    \centering
    \caption{Replication of First-Stage Estimates, Predicting State Funding by Funding Shift-Share, in IPEDS Data 2012--2019 \citep{hinrichs2022state}.}
    \makebox[\textwidth][c]{\input{tables/firststage-replicate.tex}}
    \label{tab:firststage-replicate}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the first stage OLS estimates of regressing state funding in years 2012-2019 but the 2011 base share version of the shift-share instrument, without adjusting for student population size.
    Standard errors are clustered at the state-year level, and university $+$ year fixed effects are included where noted.
    The F statistic is for the excluded instrument (not for overall regression), after accounting for fixed effects and cluster dependence \citep{olea2013robust}.
\end{table}
This definition contrasts with the analysis in \autoref{sec:empirics}, and \cite{NBERw23736}, \cite{NBERw23736}.
All three use a shift-share based in the year 1990, or averaged 1990--1993, analyse outcomes in the 1990s and 2000s, and adjust for student population size.

\begin{table}[H]
    \singlespacing
    \centering
    \caption{Replication of Second-Stage Estimates, Effects of State Funding on Faculty Counts, in IPEDS Data 2012--2019 \citep{hinrichs2022state}.}
    \makebox[\textwidth][c]{\input{tables/facultycount-replicate.tex}}
    \label{tab:facultycount-replicate}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the second stage OLS and 2SLS estimates of regression specification \eqref{eqn:secondstage}, using the state funding shift-share based in 2011 (as described above) to instrument for state funding in the columns labelled 2SLS.
    State funding is measured in \$ millions.
    Each observation is a public university-year, in the IPEDS data.
    This shows the effect of a state funding cut of \$$1,000,000$ on the number of faculty --- i.e.,
    a cut of \$$1,000,000$ leads to 0.4 fewer lecturers according to column 2.
    Outcome-mean is the mean of the outcome, the number faculty at the university.
    Standard errors are clustered at the state-year level, and university $+$ year fixed effects are included through-out.
\end{table}

According to the \cite{hinrichs2022state} specification, faculty counts are correlated by OLS measures, but the IV estimates show that there is no relationship between raw counts of state funding and faculty counts in the 2010s --- replicating the main finding in \cite{hinrichs2022state}.

On the other hand, there is a relationship between the amount of state funding per student and the number of faculty per student for decades previously.
This means that not accounting for rising student enrolment mistakes the fundamental story --- see \autoref{sec:trends}.
Furthermore, most of the substitution towards lecturers happened over the decades before 2010.
\autoref{tab:facultycount-replicated-corrected} shows that the substitution towards lecturers was much larger in the 1990s and 2000s, and indistinguishable from zero for the later period.
This both replicates the primary \cite{hinrichs2022state} finding, and shows when the substitution towards lecturers happened: in the 1990s and 2000s.
Furthermore, the trends in falling state funding for universities were most intense in the 1990s until late 2000s, mostly plateauing for 2010 and onwards.
Looking for the effects of stagnating state funding for higher education by only considering the years 2012--2019 is ignoring the time period when state funding stagnated, and when universities substituted towards lecturers.
